
[![Travis-CI Build Status](https://travis-ci.org/reconhub/outbreaker2.svg?branch=master)](https://travis-ci.org/reconhub/outbreaker2)
[![Coverage Status](https://codecov.io/github/reconhub/outbreaker2/coverage.svg?branch=master)](https://codecov.io/github/reconhub/outbreaker2?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/outbreaker2)](https://cran.r-project.org/package=outbreaker2)



*outbreaker2: a framework for reconstructing disease outbreaks*
---------------------------------------------------------------

*outbreaker2* provides a Bayesian framework for integrating epidemiological and
 genetic data to reconstruct transmission trees of densely sampled outbreaks. It
 re-implements, generalises and replaces the model of
 [*outbreaker*](https://github.com/thibautjombart/outbreaker). It uses a modular
 approach which enables fine customisation of priors, likelihoods and parameter
 movements (see [customisation
 vignette](http://www.repidemicsconsortium.org/outbreaker2/articles/customisation.html)).


<br>

This project is currently in testing stage. Please contact [the
authors](mailto:thibautjombart@gmail.com) if you plan on using it.




<br>

Installation
-------------

To install the development version from github (requires Rtools on windows and
GSL headers on all platforms):

```{r, eval = FALSE}
devtools::install_github("thibautjombart/outbreaker2")
```

To add local copies of the vignettes, you will need to specify:
```{r, eval = FALSE}
devtools::install_github("thibautjombart/outbreaker2", build_vignettes = TRUE)
```

Then, to load the package, use:

```{r, eval = FALSE}
library("outbreaker2")
```



<br>

Documentation
-------------

*outbreaker2* is fully documented on a [dedicated
 website](http://www.repidemicsconsortium.org/outbreaker2/). 

It also comes with the following vignettes:

```{r, echo = FALSE, message = FALSE}
invisible(require(printr))
```

```{r, list_vignettes}
vignette(package = "outbreaker2")
```



<br>

Contributors
------------
- [Thibaut Jombart](https://github.com/thibautjombart)
- [Finlay Campbell](https://github.com/finlaycampbell)
- [Rich Fitzjohn](https://github.com/richfitz)


See details of contributions on: <br>
https://github.com/reconhub/outbreaker2/graphs/contributors



Contributions are welcome via **pull requests**.

Please note that this project is released with a [Contributor Code of
Conduct](CONDUCT.md). By participating in this project you agree to abide by its
terms.

**Maintainer:** Thibaut Jombart (thibautjombart@gmail.com)
