---
title: "Using custom priors, likelihood, or movements in outbreaker2"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Customisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width=8, 
  fig.height=5, 
  fig.path="figs-customisation/"
)
```


In this vignette, we show how custom functions for priors, likelihood, or
movement of parameters and augmented data can be used in *outbreaker2*. In all
these functions, the process will be similar:

1. write your own function with the right arguments
2. pass this function as an argument to a `custom...` function
3. pass the result to *outbreaker2*

Note that 2-3 can be a single step if passing the function to the arguments of
*outbreaker2* directly. Also note that **all priors and likelihoods are expected
on a log scale**. Finally, also note that while the various `custom...`
functions will try to some extent to check that the provided functions are
valid, such tests are very difficult to implement. In short: you are using these
custom features at your own risks - make sure these functions work before
passing them to *outbreaker2*.




<br>

# Customising priors

Priors of *outbreaker2* must be a function of an `outbreaker_param` list (see
`?outbreaker_param`). Here, we decide to use a step function rather than the
default Beta function as a prior for *pi*, the reporting probability.


We start by defining two functions: an auxiliary function `f` which returns
values on the natural scale, and which we can use for plotting the prior
distribution, and then a function `f_pi` which will be used for the
customisation.

```{r, f_pi}
f <- function(pi) {
    ifelse(pi < 0.8, 0, 5)
}

f_pi <- function(param) { 
    log(f(param$pi))
}

plot(f, type = "s", col = "blue", 
     xlab = expression(pi), ylab = expression(p(pi)), 
     main = expression(paste("New prior for ", pi)))

```


While `f` is a useful function to visualise the prior, `f_pi` is the function
which will be passed to `outbreaker`. To do so, we pass it to `custom_priors`:

```{r, custom_prior}

priors <- custom_priors(pi = f_pi)
priors

```

Note that `custom_priors` does more than just adding the custom function to a
list. For instance, the following customisations are all wrong, and rightfully
rejected:

```{r, wrong_prior}

## wrong: not a function
## should be pi = function(x){0.0}
custom_priors(pi = 0.0)

custom_priors(pi = function(x, y){0.0})

```



