---
title: "Using custom priors, likelihood, or movements in outbreaker2"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Customisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width=8, 
  fig.height=5, 
  fig.path="figs-customisation/"
)
```


In this vignette, we show how custom functions for priors, likelihood, or movement of parameters and augmented data can be used in *outbreaker2*. In all these functions, the process will be similar:

1. write your own function with the right arguments
2. pass this function as an argument to a `custom...` function
3. pass the result to *outbreaker2*

Note that 2-3 can be a single step if passing the function to the arguments of *outbreaker2* directly. Also note that **all priors and likelihoods are expected on a log scale**. Finally, also note that while the various `custom...` functions will try to some extent to check that the provided functions are valid, such tests are very difficult to implement. In short: you are using these custom features at your own risks - make sure these functions work before passing them to *outbreaker2*.




<br>

# Customising priors

Priors of *outbreaker2* must be a function of an `outbreaker_param` object.


```{r, data}



```



