---
title: "Introduction to outbreaker2"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width=8, 
  fig.height=5, 
  fig.path="figs-introduction/"
)
```


This tutorial provides a worked example of outbreak reconstruction using
*outbreaker2*. For installation guidelines, a general overview of the package's
functionalities as well as other resources, see the 'overview' vignette:
```{r, eval=FALSE}
vignette("Overview", package = "outbreaker2")
```

We will be analysing a small simulated outbreak distributed with the package,
`fake_outbreak`. This dataset contains simulated dates of onsets and pathogen
genome sequences for 30 cases:


```{r, data}
library(outbreaker2)
library(ape)

fake_outbreak
```




<br>

# Running the analysis with defaults



<br>

# Customising settings and priors




<br>

# Analysing the results

## Plotting results

## Using `summary`

## Consensus tree


