---
title: "outbreaker2: Rcpp API"
author: "Thibaut Jombart"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Rcpp API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



# List of available functions

Here are all the C++ functions bound to R via Rcpp available in *outbreaker2*:

```{r}

env <- asNamespace("outbreaker2")
list_cpp_functions <- sort(ls(envir = env, pattern = "^cpp"))
list_cpp_functions

```



# Function calls

These functions take the following arguments:
```{r, arguments}

get_args <- function(f) args(get(f, envir = env))
get_formals <- function(f) formals(get(f, envir = env))

list_args <- lapply(list_cpp_functions, get_args)
names(list_args) <- list_cpp_functions
list_args

```




# Arguments

These arguments are:
```{r}

list_formals <- lapply(list_cpp_functions, get_formals)
names(list_formals) <- list_cpp_functions
args <- sort(unique(unlist(lapply(list_formals, names))))

```

- **`alpha`**: a vector of integers of length 'N' (number of cases), indicating infectors of each case, with values from 1 to N; missing values should be `NA`

- **`config`**: a list containing configuration settings as returned by `create_config`

- **`custom_function`**: a R function for a custom prior, with a single argument, which must be a list of parameters and augmented data with the class `outbreaker_param`; returned values must be **on the log scale**

- **`custom_functions`**: a list of R functions obeying the rules of `custom_function`, named according to the priors; currently available names are:
```{r, echo = FALSE}
temp <- sub("cpp_prior_", "",
     	    ls(envir = env, pattern = "^cpp_prior"))
setdiff(temp, "all")	    
```

- **`custom_ll`**: a R function for a custom likelihood, taking two arguments: `data` (see `data`), and `param` (see `param`)

- **`custom_prior`**: same as `custom_function`

- **`data`**: a valid 'outbreaker_data' list

- **`i`**: an integer scalar indicating the index of a case, from 1 to N (number of cases) 

- **`list_custom_ll`**: a list of R functions obeying the rules of `custom_ll`, named according to the computed likelihood component; available names are: 
```{r, echo = FALSE}
temp <- sub("cpp_ll_", "",
     	    ls(envir = env, pattern = "^cpp_ll"))
setdiff(temp, c("timing", "all"))	    
```

- **`param`**: a list containing parameters and augmented data with the class `outbreaker_param`

- **`t_inf`**: a vector of integers of length N (number of cases), indicating infection dates of each case; missing values should be `NA`


- **`x`**: a vector of integers to be sampled from
